\begin{abstract}
The rapid proliferation of the Internet of Things (IoT) has fundamentally expanded the attack surface of modern networks, necessitating the deployment of Intrusion Detection Systems (IDS) directly at the computational edge. However, transitioning from centralized cloud architectures to edge environments presents a critical paradox: highly accurate Deep Learning models are notoriously "black-boxes" that lack operational transparency, and their extreme computational complexity renders them incompatible with resource-constrained IoT gateways like the Raspberry Pi. This opacity breeds a profound lack of trust among security administrators, who require rapid, actionable intelligence rather than opaque binary alerts. To resolve this dilemma, this paper proposes a comprehensively Trustworthy Edge Security framework. We design a real-time, distributed streaming architecture utilizing Apache Kafka to ingest and process network telemetry directly on Raspberry Pi edge nodes. To overcome the computational and interpretative bottlenecks, we deploy a lightweight, highly optimized machine learning model augmented with Explainable AI (XAI) mechanisms. Specifically, we utilize SHapley Additive exPlanations (SHAP) both as a pre-deployment feature-pruning strategy—drastically reducing processing overhead—and as a real-time interpretability engine. Evaluated on the benchmark CICIDS2017 dataset, our empirical results demonstrate that the SHAP-optimized edge model maintains exceptionally high threat-detection accuracy while dramatically minimizing memory consumption and inference latency on the Raspberry Pi. By explicitly mapping the contribution of individual network features to every generated alert, our framework transforms the opaque IDS into a transparent, trustworthy, and resource-efficient sentinel capable of securing the modern IoT ecosystem.
\end{abstract}
